\bf Hierarchical Network Structure Promotes Dynamical Robustness

### 2015-07-22 tex/introduction.tex
add section corresponding to introduction
move reference to total ordering from the introduction to the quantification of network hierarchy section

%!TEX root = ../paper.tex
The traditional approach taken in the study of chemical reaction, gene-regulatory, population, and ecosystem networks is to derive a system of differential equations to model a particular biological network, attempt to fit that model to data and adjust the modeling assumptions along with parameter values until a good fit is achieved \cite{Meyer2014}. Over evolutionary timescales, one expects to observe changes in the model of best fit. All of these models utilize essentially equivalent mathematical structures, instances of which are sampled in the evolutionary process (\ref{fig:biomodelexamples}, \cite{RossCr2003,Palsson2011a,Sauro2012}). Developing unified mathematical descriptions of each of these that can be embedded into models of the evolutionary process is one of the paramount goals of systems biology.
% \cite{RossCr2003,Alon2006,Palsson2006,HamidBolouri2008,Palsson2011a,Voit2012,Sauro2012}

@@

Here we determine the relationship between network hierarchy, a topological property, and \emph{robustness}, a dynamical property. Robustness is of interest in biological systems at all scales, and has been previously studied in the context of biochemical networks \cite{Alon1999,Shinar2010}, gene-regulatory networks \cite{VanNimwegen1999,Siegal2002,Draghi2010,Wagner2013}, and ecological networks \cite{Rohr2014}. Over physiological timescales, the robustness of a particular network state may be evaluated by determining its linear stability \cite{Davis1962}. Network states that are linearly stable, are robust to perturbations in the states of their components. For example, in the case of gene-regulatory networks, a state vector containing protein molecule counts or concentrations of each gene that is stable with respect to the dynamics of the network will exponentially suppress any relatively small modifications to the state of a gene and return to the initial stable state.

@@

We demonstrate that systems exhibiting maximal robustness with respect to this definition have the most hierarchical network topology and explain why this results from an invariance of robustness to particular kinds of transformations of the network topology. For the purpose of formulating this result, the maximally hierarchical network is considered to be the graph associated to the total ordering (\refsupp{} and \cite{Cormen2009}). An example of this for three system components is shown in \reffigscc{} top. We use a measure of hierarchy based upon the edit distance from this maximally hierarchical network \cite{Axenovich2011}. Our results hold for networks of arbitrary size and are independent of the probability distribution from which the strengths of interaction are sampled.
% \cite{WADDINGTON1942a,VanNimwegen1999,Siegal2002,Ciliberti2007b,Ciliberti2007,Wagner2013}

>>>>

%!TEX root = ../paper.tex
\section{Introduction}

The traditional approach taken in the study of chemical reaction, gene-regulatory, population, and ecosystem networks is to derive a system of differential equations to model a particular biological network, attempt to fit that model to data and adjust the modeling assumptions along with parameter values until a good fit is achieved \cite{Meyer2014}. Over evolutionary timescales, one expects to observe changes in the model of best fit. All of these models utilize essentially equivalent mathematical structures, instances of which are sampled in the evolutionary process (\ref{fig:biomodelexamples}, \cite{RossCr2003,Palsson2011a,Sauro2012}). Developing unified mathematical descriptions of each of these that can be embedded into models of the evolutionary process is one of the paramount goals of systems biology.
% \cite{RossCr2003,Alon2006,Palsson2006,HamidBolouri2008,Palsson2011a,Voit2012,Sauro2012}

@@

Here we determine the relationship between network hierarchy, a topological property, and \emph{dynamical robustness}, a property of the dynamics. Robustness is of interest in biological systems at all scales, and has been previously studied in the context of biochemical networks \cite{Alon1999,Shinar2010}, gene-regulatory networks \cite{VanNimwegen1999,Siegal2002,Draghi2010,Wagner2013}, and ecological networks \cite{Rohr2014}. Over physiological timescales, the robustness of a particular network state may be evaluated by determining its linear stability \cite{Davis1962}. Network states that are linearly stable, are robust to perturbations in the states of their components. For example, in the case of gene-regulatory networks, a state vector containing protein molecule counts or concentrations of each gene that is stable with respect to the dynamics of the network will exponentially suppress any relatively small modifications to the state of a gene and return to the initial stable state.

@@

We demonstrate that systems exhibiting maximal robustness with respect to this definition have the most hierarchical network topology relative to a specification of system modules and explain why this results from an invariance of robustness to particular kinds of transformations of the network topology. Our results hold for networks of arbitrary size and are independent of the probability distribution from which the strengths of interaction are sampled.
% \cite{WADDINGTON1942a,VanNimwegen1999,Siegal2002,Ciliberti2007b,Ciliberti2007,Wagner2013}

### 2015-07-22 tex/overview.tex
move reference to total ordering from the introduction to the quantification of network hierarchy section

The relationship between $G$ and $\hier(G)$ for all $G$ with a given number of vertices suggests a heuristic method of quantifying the degree of hierarchy of a given graph and thus of the system structure it represents. The most hierarchical system is considered to be the graph corresponding to the total ordering, which for three nodes is given in \reffigscc{} (top panel). This graph maximizes the number of links between strongly connected components, which also implies maximizing the number of strongly connected components. The graph edit distance (ED) on a fixed number of vertices from one graph to another is defined as the minimum number of modifications of the first graph in order to transform it into the second \cite{Axenovich2011}. This distance between any given graph and the total ordering thus quantitatively represents how far a graph is from being maximally hierarchical. In this work we take $max(ED) - ED$ to be the definition of hierarchy, where $max(ED)$ is the maximum edit distance for all graphs with a given number of nodes.

>>>>

The relationship between $G$ and $\hier(G)$ for all $G$ with a given number of vertices suggests a heuristic method of quantifying the degree of hierarchy of a given graph and thus of the system structure it represents. The most hierarchical system is considered to be the graph corresponding to the total ordering (\refsupp{} and \cite{Cormen2009}), which for three nodes is given in \reffigscc{} (top panel). This graph maximizes the number of links between strongly connected components, which also implies maximizing the number of strongly connected components. The graph edit distance (ED) on a fixed number of vertices from one graph to another is defined as the minimum number of modifications of the first graph in order to transform it into the second \cite{Axenovich2011}. This distance between any given graph and the total ordering thus quantitatively represents how far a graph is from being maximally hierarchical. In this work we take $max(ED) - ED$ to be the definition of hierarchy, where $max(ED)$ is the maximum edit distance for all graphs with a given number of nodes.
